---
title: "Exploratory Data Analysis (EDA)"
output:
  md_document:
    variant: markdown_mmd
  html_document:
    df_print: paged
---

```{r setup, message=FALSE}
library(MASS) # Needed for `birthwt` dataset
library(tidyverse)

# Non-tidyverse packages
library(skimr) # This appears to be tidyverse-adjacent
library(summarytools)

df <- birthwt
df$race_factor <- factor(df$race)
```

## Initial exploration of a new dataset

### `glimpse()` lists out columns, their types, and some example values

This function is [part of `tibble`](https://tibble.tidyverse.org/reference/glimpse.html), the tidyverse extension of `data.frame`.

```{r}
df %>% glimpse()
```

### `summary()` shows summary stats

This is part of base R and is a convenient, if not overly pretty, way of getting summary stats like mean, median, etc.

```{r}
df %>% summary()
```

### `table` to show frequency counts for a variable

Also part of base R, this is another quick way to show frequency counts.

```{r}
df$race %>% table()
```



### `skimr`

This is a tidy-compatible package for summarizing data.

```{r}
df %>% skimr::skim()
```

Here's what it looks like with a string variable (`Species`):

```{r}
iris %>% skimr::skim()
```

## Another option: `summarytools::dfSummary()`

This is similar to `skimr::skim()`, but is less conservative about horizontal space and can in theory use image-based graphs rather than ASCII for histograms (though this requires X11, which doesn't work on my system for some reason):

```{r}
dfSummary(iris)
```



## Frequency counts of categorical variables

When using R, I frequently miss Stata's `tab` function for summarizing categorical data. This is what it looks like:

```
. webuse iris
(Iris data)

. tab iris

       Iris |
    species |      Freq.     Percent        Cum.
------------+-----------------------------------
     setosa |         50       33.33       33.33
 versicolor |         50       33.33       66.67
  virginica |         50       33.33      100.00
------------+-----------------------------------
      Total |        150      100.00
```

To achieve this in R, the best thing I've found is [`summarytools::freq()`](https://github.com/dcomtois/summarytools):

```{r}
freq(iris$Species)
```

## Contingency tables ("2x2 tables")

I also miss `tab2` from Stata, which looks like this:

```
. tab tobacco parent, col

+-------------------+
| Key               |
|-------------------|
|     frequency     |
| column percentage |
+-------------------+

                      |   1 = either parent
                      |        smoked
        tobacco usage | nonsmokin    smoking |     Total
----------------------+----------------------+----------
         0 cigarettes |     5,177      4,292 |     9,469 
                      |     70.49      56.06 |     63.13 
----------------------+----------------------+----------
1 to 7 cigarettes/day |     1,593      2,213 |     3,806 
                      |     21.69      28.91 |     25.37 
----------------------+----------------------+----------
8 to 12 cigarettes/da |       378        672 |     1,050 
                      |      5.15       8.78 |      7.00 
----------------------+----------------------+----------
more than 12 cigarett |       196        479 |       675 
                      |      2.67       6.26 |      4.50 
----------------------+----------------------+----------
                Total |     7,344      7,656 |    15,000 
                      |    100.00     100.00 |    100.00 
```

You can accomplish something similar with `summarytools::ctable()`

```{r}
summarytools::ctable(tobacco$smoker, tobacco$diseased, prop = "c")
```

## Other resources

- [Parts 1](https://www.littlemissdata.com/blog/simple-eda) and [2](https://www.littlemissdata.com/blog/simple-eda) on exploratory data anlaysis from Little Miss Data.
