---
title: "Data Manipulation"
output:
  md_document:
    variant: markdown_mmd
  html_document:
    df_print: paged
---

## Setup

```{r setup, message=FALSE}
library(MASS) # Needed for `birthwt` dataset
library(tidyverse)

df <- birthwt
```

## Aggregating

### Maximum value across rows

```{r}
df %>%
  select(age, lwt) %>%
  mutate(age = age * 5, test = 110) %>% # Make `age` column large enough to sometimes be > `lwt`
  mutate(
    row_max = pmax(age, lwt, test)
  ) %>% 
  head()
```


## Filtering

### Select rows where variable contains string

`str_detect()` allows you to filter strings with a regular expression. In this example, I use `a$`, which looks for the letter `a` at the end of each value for `Species`. This has the effect of filtering out all the rows with `Species == "versicolor"`.

```{r}
iris %>% 
  filter(str_detect(Species, 'a$')) %>% 
  select(Species) %>% 
  table()
```

## Modifying

### Update a subset of rows based on criteria

#### Using `mutate()`

The simplest way is to use `mutate()` and `case_when()`.

```{r}
df_example <- tibble(x = 1:5, y = c('a', 'b', 'c', 'd', 'e'))

# A tibble: 5 x 2
#       x y    
#   <int> <chr>
# 1     1 a    
# 2     2 b    
# 3     3 c    
# 4     4 d    
# 5     5 e 

df_example <- df_example %>% 
  mutate(
    y = case_when(
      x > 3 ~ 'hello',
      TRUE ~ y # Without this, `y` will be `NA` for the first 3 rows
    )
  )

df_example
```

You can also use `if_else()` to achieve a similar effect as `case_when()` if you don't need multiple cases.


#### Using `mutate_at()`

If you want to use `mutate_at()` the syntax is somewhat different:

```{r}
df_example <- tibble(x = 1:5, y = c('a', 'b', 'c', 'd', 'e'))

# A tibble: 5 x 2
#       x y    
#   <int> <chr>
# 1     1 a    
# 2     2 b    
# 3     3 c    
# 4     4 d    
# 5     5 e 

df_example %>% 
  mutate_at(
    vars(x),
    list(~ case_when(
      . == 2 ~ 0L,
      TRUE ~ .
    ))
  )
```

Note that you have to use `. == 2 ~ 0L` or `TRUE ~ as.numeric(.)` in this example to avoid a type error. I have no idea why R can't figure this out for itself, but it apparently can't.

#### Masks

You can also use a mask:

```{r}
df_example <- tibble(x = 1:5, y = c('a', 'b', 'c', 'd', 'e'))

# A tibble: 5 x 2
#       x y    
#   <int> <chr>
# 1     1 a    
# 2     2 b    
# 3     3 c    
# 4     4 d    
# 5     5 e 

mask <- df_example$x > 3
df_example[mask, 'y'] <- c('hello', 'goodbye')

df_example
```


### Using other columns in `mutate()`, `mutate_at()`, or `mutate_all()`

```{r}
df_example <- tibble(x=1:5, y=10:14, z=100:104)
df_example
```

This is simple for `mutate()`:

```{r}
df_example %>% 
  mutate(
    new_col = x + y
  )
```

For `mutate_at()` and `mutate_all()`, it is a bit more complex:

```{r}
df_example %>% 
  mutate_at(
    vars(x, y),
    ~ . + z
  )
```

This complexity is because `mutate_at()` requires that you provide a function rather than simply providing an expression like `new_col = x + y`.

The `~` character [is used in base R with formulas](https://stackoverflow.com/questions/56650810/what-does-the-tilde-mean-in-this-context-of-r-code), but is overloaded in `purrr` (one of the `tidyverse` packages) to implement anonymous functions (i.e. a function that's not assigned to a name).

This is functinally equivalent to `~ . + z` but with more verbose syntax:

```{r}
f <- function(., other_col) . + other_col
df_example %>% 
  mutate_at(
    vars(x, y),
     f, other_col = quote(z)
  )
```

The `quote()` function makes this somewhat confusing, and a technical explanation is beyond the scope of this guide. However, you can [read more about quoting in R here](https://tidyeval.tidyverse.org/sec-up-to-speed.html).




## Miscellaneous

- Convert column to vector: use `deframe()` with a Tibble.